Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality

### 2021-10-04 Code/Manuscript-SMOTE.Rmd
analyses complete

---
title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
author: "Sarah J. Hart"
email: "sarah.hart@colostate.edu"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: 
  bookdown::word_document2:
    reference_docx: WordTemplate.docx
    fig_caption: yes
    toc: no
    number_sections: no
    df_print: kable
csl: "`r here:::here('ecology.csl')`"
bibliography: "`r here:::here('MyLib.bib')`"
urlcolor: blue
linkcolor: blue
citationcolor: blue
---

@@

### Import libraries
ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'themis', 'foreach'))

#foreach', 'parallel', 'doParallel', 'quickPlot', 'reshape2','vegan', 'FSA', 
# Set seed
set.seed(2020)

@@

Paragraph 2: Theoretical context - diversity & natural enemies
- Many studies have sought to understand how community diversity influences interactions between natural enemies and their resources.
- Key hypotheses:
1 - associational resistance - the abundance or damage of herbivores is lower when plant diversity is greater [@barbosa_associational_2009, @jactel_tree_2021]
hypotheses:
1 - resource concentration: the presence of heterospecific neighbors around a tree of a focal species grown in mixed stands leads to a lower probability of host tree finding by insect herbivores due to lower host abundance or frequency [@root_organization_1973]
2 - host apparency: 
3 - decoy: preference for non-host trees

However, benefits provided by mixtures are less evident for larger-scale disturbances [@jactel_tree_2021]

suggesting that changes in the structure of host communities, rather than biodiversity per se, can explain when a dilution effect should be observed.

Most studies have focused on individual- or population-level outcomes.

@@

Bark beetles (Curculionidae: Scolytinae) are among the few native insect species that can kill large numbers of trees in a single year. Bark beetles bore through the bark, where they mate and oviposit their eggs. Concurrently, bark beetles introduce pathogenic fungi Larvae feeding upon the phloem and fungal spread stop the translocation of water and nutrients and cause tree death. Conifer defense against bark beetles consists primarily of resin exudation that physically expels the beetle and allelochemicals, which repel or kill beetles. To overcome these defenses and colonize live trees, bark beetles rely on a mass-attack strategy, where pioneering beetles emit aggregation pheromones that call conspecifics to the tree. Typically bark beetles exist at low population levels and attack weakened trees, but as populations increase bark beetles attack increasingly better defended trees.For instance, in the continental western United States bark beetles have killed more than XXXX trees over the past X decades. Such severe mortality occurs only during outbreaks when pheromone-mediated mass-attack allows bark beetles to overcome tree defenses. 
For instance, a primary goal of the Western Bark Beetle is to promote resilience of forests to bark beetle outbreaks by increasing the diversity of age classes and tree species.

If greater host richness increases niche space for bark beetles, then community-level occurrence of bark beetle activity will be greater in more diverse communities. community-level severity, here the amount and proportion of basal area affected by bark beetles, will be lower in stands with more host species.

To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/). Specifically we ask: (1) Is the occurrence or severity of bark beetle

@@

The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming; Fig. \@ref(fig:FigSA)

@@

df$agents <-  do.call(paste, c(df[c("agentSB", "agentMPB", "agentWBBB")],sep="-"))
df$agents <- factor(df$agents, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"))

df$nagents <- df$agents
df$nagents <- factor(df$nagents, levels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"), labels=c(0, 1, 1, 2, 1, 2, 2, 3))

df$hostsPICO <- ifelse(df$TPH.PICO>0, "PICO", "absent")
df$hostsPIEN <- ifelse(df$TPH.PIEN>0, "PIEN", "absent")
df$hostsABLA <- ifelse(df$TPH.ABLA>0, "ABLA", "absent")
df$hosts <-  do.call(paste, c(df[c("hostsPICO", "hostsPIEN", "hostsABLA")],sep="-"))
df$hosts <- factor(df$hosts, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA', 'absent-absent-absent'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA', 'none'), ordered=TRUE)

df$nhosts <- df$hosts
df$nhosts <- factor(df$nhosts, levels=c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA', 'none'), labels=c(1, 1,1,2,2,2,3,0))

@@

### Descriptive analyses
```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-df-merge-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), ordered=T)
df$hostsFPlong <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

df$nagentsF <- factor(df$nagents, levels=0:3, ordered=T)
df$agentsF <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB & SB","MPB & WBBB",  "SB & WBBB", "MPB, SB & WBBB"), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB\n& SB","MPB &\nWBBB",  "SB &\nWBBB", "MPB,\nSB, &\nWBBB"), ordered=T)
df$agentsFplong <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB & SB","MPB & WBBB",  "SB & WBBB", "MPB, SB & WBBB"), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB","MPB & WBBB",  "SB & WBBB", "MPB, SB, & WBBB"), ordered=T)

@@

```{r map study area of hosts, eval=F}
f19 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf"), proxy=F)
f93 <- stars::read_stars(here("/Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf"), proxy=F)
f108 <- stars::read_stars(here("/Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf"), proxy=F)

>>>>

---
title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
author: "Sarah J. Hart et al."
email: "sarah.hart@colostate.edu"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: 
  bookdown::word_document2:
    reference_docx: WordTemplate.docx
    fig_caption: yes
    toc: no
    number_sections: no
    df_print: kable
csl: "`r here:::here('ecology.csl')`"
bibliography: "`r here:::here('MyLib.bib')`"
urlcolor: blue
linkcolor: blue
citationcolor: blue
editor_options: 
  chunk_output_type: inline
---

@@

### Import libraries
ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'themis', 'foreach',  'FSA', 'parallel'))

# Set seed
set.seed(2020)

# Set number of cores for parralell processing
cores <- parallel::detectCores()

@@

Paragraph 2: Theoretical context - diversity & natural enemies

Many studies have sought to understand how community diversity influences interactions between natural enemies and their resources. In the case of plant-herbivore interactions, the greater plant diversity often reduces herbivory effects on plants at both the individual and population level [i.e. 'associational resistance, @barbosa_associational_2009, @jactel_tree_2021]. The resource concentration hypothesis suggests that specialists insect herbivores are less likely to find and stay in stands where their hosts are less abundant [@root_organization_1973]. This may occur if insects are less likely to find a host tree due [i.e. 'host dilution effect', @] or chemical masking of the focal tree by heterospecific neighbors [i.e. 'host apparency hypothesis', @Castagneryrol_effects_2014]

[i.e., 'resource concentration hypothesis', @root_organization_1973].

However, benefits provided by mixtures are less evident for larger-scale disturbances [@jactel_tree_2021]
suggesting that changes in the structure of host communities, rather than biodiversity per se, can explain when a dilution effect should be observed.

1 - associational resistance - the abundance or damage of herbivores is lower when plant diversity is greater [@barbosa_associational_2009, @jactel_tree_2021]
hypotheses:
1 - resource concentration: the presence of heterospecific neighbors around a tree of a focal species grown in mixed stands leads to a lower probability of host tree finding by insect herbivores due to lower host abundance or frequency [@root_organization_1973]
2 - host apparency: 
3 - decoy: preference for non-host trees

H
Most studies have focused on individual- or population-level outcomes.

@@

Bark beetles (Curculionidae: Scolytinae) are among the few native insect species that can kill large numbers of trees in a single year. Bark beetles bore through the bark, where they mate and oviposit their eggs. Concurrently, bark beetles introduce pathogenic fungi Larvae feeding upon the phloem and fungal spread stop the translocation of water and nutrients and cause tree death. Conifer defense against bark beetles consists primarily of resin exudation that physically expels the beetle and allelochemicals, which repel or kill beetles. To overcome these defenses and colonize live trees, bark beetles rely on a mass-attack strategy, where pioneering beetles emit aggregation pheromones that call conspecifics to the tree. Typically bark beetles exist at low population levels and attack weakened trees, but as populations increase bark beetles attack increasingly better defended trees.For instance, in the continental western United States bark beetles have killed more than XXXX trees over the past X decades. Such severe mortality occurs only during outbreaks when pheromone-mediated mass-attack allows bark beetles to overcome tree defenses.

For instance, a primary goal of the Western Bark Beetle is to promote resilience of forests to bark beetle outbreaks by increasing the diversity of age classes and tree species.

To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/).

We expect that co-occurrence of trees of different host species will be common, but co-occurrence of stand conditions suitable for multiple bark beetle species will not. Given stands where conditions are suitable for multiple bark beetle species, occurrence of bark beetle-driven tree mortality will be greater (i.e. diversity begets diversity), but severity will be lower (i.e. resource concentration).

We also expected the severity of bark beetle infestation to vary with the number of agents present, but with two alternative hypotheses. If the population dynamics of each bark beetle species are independent, then stands with multiple agents will experience higher tree mortality than stands with only one agent (i.e. additive effects). Alternatively, lower tree mortality may occur if concurrent outbreaks of bark beetles of different species cause semiochemical confusion or if competitive release increases tree defensive capacity.

@@

The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming)

@@

df$agents <-  do.call(paste, c(df[c("agentSB", "agentMPB", "agentWBBB")],sep="-"))
df$agents <- factor(df$agents, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"))

df$nagents <- df$agents
df$nagents <- factor(df$nagents, levels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"), labels=c(0, 1, 1, 2, 1, 2, 2, 3))

df$hostsPICO <- ifelse(df$TPH.PICO>0, "PICO", "absent")
df$hostsPIEN <- ifelse(df$TPH.PIEN>0, "PIEN", "absent")
df$hostsABLA <- ifelse(df$TPH.ABLA>0, "ABLA", "absent")
df$hosts <-  do.call(paste, c(df[c("hostsPICO", "hostsPIEN", "hostsABLA")],sep="-"))
df$hosts <- factor(df$hosts, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA', 'absent-absent-absent'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), ordered=TRUE)

df$nhosts <- df$hosts
df$nhosts <- factor(df$nhosts, levels=c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), labels=c(1, 1,1,2,2,2,3,0))

@@

```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-df-merge-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hostsFP <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

df$nagentsF <- factor(df$nagents, levels=0:3, ordered=T)
df$agentsF <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB","MPB+WBBB",  "SB+WBBB", "MPB+SB+WBBB"), ordered=T)

@@

### Model the suitable stand conditions
Construct random forest models using synthetic minority oversampling technique (SMOTE), an approach for dealing with imbalanced data. The SMOTE approach oversamples the minority class by synthesizing new cases from the minority class. We compared this approach with two other common approaches for dealing with imbalanced data in a random forest framework - balanced and weighted random forests and selected the approach that classified affected plots with the greatest accuracy.

```{r fit MPB RF model SMOTE, eval=F}
response <- "agentMPB"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO") # predictor variables
datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit) # create object with training data
datz.test <- testing(splitit) # create object with testing data
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases

# set sample weights (not used in SMOTE approach)
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")

# tune model
tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))

tuneres <- rf_tune_results %>% collect_metrics() # collect metrics

param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters

# Use backward selection to identify important variables
#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
#registerDoSEQ()
#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
vars.final <- vars
BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)

rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set. 
test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)

write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)

# generate predictions for entire dataset
x <- datBB.c$PLT_CN
x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
colnames(x) <- c("PLT_CN", "SBpred")
write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)

x <- as.data.frame(ranger_obj$variable.importance)
write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```

```{r fit MPB RF model Balanced, eval=F}
response <- "agentMPB"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts) %>% set_mode("classification")

# tune model
tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))

tuneres <- rf_tune_results %>% collect_metrics() # collect metrics

param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters

rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set. 
test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED.csv")), row.names=F)

write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED.csv")), row.names=F)

# generate predictions for entire dataset
x <- datBB.c$PLT_CN
x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
colnames(x) <- c("PLT_CN", "SBpred")
write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED.csv")), row.names=F)

x <- as.data.frame(ranger_obj$variable.importance)
write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED.csv")), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED.csv")), row.names=F)
```

```{r fit MPB RF model Weighted, eval=F}
response <- "agentMPB"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")

# tune model
tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))

tuneres <- rf_tune_results %>% collect_metrics() # collect metrics

param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters

rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set. 
test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED.csv")), row.names=F)

write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED.csv")), row.names=F)

# generate predictions for entire dataset
x <- datBB.c$PLT_CN
x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
colnames(x) <- c("PLT_CN", "SBpred")
write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED.csv")), row.names=F)

x <- as.data.frame(ranger_obj$variable.importance)
write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED.csv")), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
```

```{r fit SB RF model SMOTE, eval=F}
response <- "agentSB"
vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")

# tune model
tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))

tuneres <- rf_tune_results %>% collect_metrics() # collect metrics

param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters

# Use backward selection to identify important variables
#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
#registerDoSEQ()
#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
vars.final <- vars
BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)

rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set. 
test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)

write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)

# generate predictions for entire dataset
x <- datBB.c$PLT_CN
x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
colnames(x) <- c("PLT_CN", "SBpred")
write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)

x <- as.data.frame(ranger_obj$variable.importance)
write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```

```{r fit SB RF model Balanced, eval=F}
response <- "agentSB"
vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")

# tune model
tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))

tuneres <- rf_tune_results %>% collect_metrics() # collect metrics

param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters

rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set. 
test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED.csv")), row.names=F)

write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED.csv")), row.names=F)

# generate predictions for entire dataset
x <- datBB.c$PLT_CN
x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
colnames(x) <- c("PLT_CN", "SBpred")
write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED.csv")), row.names=F)

x <- as.data.frame(ranger_obj$variable.importance)
write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED.csv")), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED.csv")), row.names=F)
```

```{r fit SB RF model Weighted, eval=F}
response <- "agentSB"
vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")

datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set model
rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")

# tune model
tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))

tuneres <- rf_tune_results %>% collect_metrics() # collect metrics

param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
